---
title: "Terrestrial Soil Carbon — Steps 1 & 2"
subtitle: "Non-Spatial Analysis and Simple Extrapolation"
date: today
engine: knitr
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
    theme: cosmo
    fig-width: 8
    fig-height: 5
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
if (!file.exists("_targets")) setwd("..")
library(targets); library(dplyr); library(ggplot2); library(knitr)

cfg                 <- tar_read(cfg)
cores_raw           <- tar_read(cores_raw)
eda_plots           <- tar_read(eda_plots)
stratum_summary     <- tar_read(stratum_summary)
step2_extrapolation <- tar_read(step2_extrapolation)
step2_strata_map    <- tar_read(step2_strata_map)
harmonized_eda      <- tar_read(harmonized_eda)
```

## Project overview

```{r}
tibble(
  Metric = c("Project name", "Scenario", "Monitoring year",
             "Total cores", "Total samples"),
  Value  = c(cfg$PROJECT_NAME, cfg$PROJECT_SCENARIO,
             as.character(cfg$MONITORING_YEAR),
             as.character(n_distinct(cores_raw$core_id)),
             as.character(nrow(cores_raw)))
) |> kable(col.names = c("", ""), align = c("l", "r"))
```

## Core locations

```{r fig.cap="Core locations coloured by stratum."}
eda_plots$spatial_map
```

## SOC profiles

```{r fig.cap="Raw SOC depth profiles. Each line is one core."}
eda_plots$depth_profiles
```

## Carbon stocks (raw)

```{r fig.cap="Total raw carbon stock per core, summed across measured depth intervals."}
eda_plots$carbon_stocks
```

## Depth harmonization

### Method

Profiles harmonized to standard depths (0–15, 15–30, 30–60, 60–100 cm) using a hybrid approach:

- **Interpolation** (within measured range): monotone Hermite spline (`monoH.FC`).
- **Extrapolation** (beyond measured range): exponential decay if Spearman ρ < −0.3
  and n ≥ 3 per core; constant otherwise. Capped at 2.5× maximum measured depth.

Target depth midpoints: `r paste(cfg$DEPTH_MIDPOINTS, "cm", collapse = ", ")`.

### Results

```{r}
stratum_summary |>
  mutate(across(where(is.numeric), \(x) round(x, 2))) |>
  select(
    Stratum              = stratum,
    `Depth (cm)`         = depth_cm_midpoint,
    `n cores`            = n_cores,
    `Mean stock (kg/m²)` = mean_stock,
    `SD (kg/m²)`         = sd_stock,
    `Mean SOC (g/kg)`    = mean_soc,
    `Mean BD (g/cm³)`    = mean_bd,
    `% extrapolated`     = pct_extrapolated
  ) |>
  kable(caption = "Harmonized carbon stocks at the standard depth intervals.")
```

## Harmonized carbon stocks

The same views as the raw data above, now computed from the depth-harmonized
profiles — same colours and style, so they can be reported alongside the raw
figures.

```{r fig.cap="Harmonized carbon stock by depth. Each line is one core, faceted by stratum.", fig.width=8, fig.height=5}
harmonized_eda$depth_profiles
```

```{r fig.cap="Total harmonized carbon stock (0–100 cm) per core, by stratum.", fig.width=8, fig.height=5}
harmonized_eda$carbon_stocks
```

## Step 2 — Simple extrapolation

Per-stratum carbon **density** (kg C/m² and Mg C/ha), summed across the
standard depth intervals. Absolute **total stock** columns (`total_stock_kg`,
`total_stock_MgC`) appear only when both `AOI_FILE` and `AOI_STRATUM_FIELD` are
set in `soil_carbon_config.R` and the polygon's stratum values match your cores.

```{r}
step2_extrapolation |>
  mutate(across(where(is.numeric), \(x) round(x, 2))) |>
  kable(caption = "Per-stratum carbon stock summary.")
```

## Thematic stock maps

A simple spatial view of the carbon stocks that needs **no** remote-sensing
covariates. Where the area of interest provides per-stratum polygons, each is
filled with its mean stock (kg C/m²); otherwise the sampling plots are shown
over the AOI, coloured by their carbon stock.

```{r}
#| fig-cap: "Carbon stock across the area of interest."
#| fig-width: 9
#| fig-height: 5
m <- step2_strata_map
if (is.null(m$full_profile)) {
  cat("No thematic map could be produced — it needs either per-stratum AOI",
      "polygons (set `AOI_STRATUM_FIELD`) or plot coordinates.")
} else {
  print(m$topsoil)
  print(m$full_profile)
}
```

## Saved output files

Every figure and table in this report is also written to `outputs/non-spatial/`
(`figures/` as PNG, `tables/` as CSV) each time the pipeline runs, so partners
can reuse them directly without rendering the report.

---

*Rendered: `r format(Sys.time(), "%Y-%m-%d %H:%M")` |
See `reports/step3_random_forest.html` for spatial predictions.*
