---
title: "Blue Carbon Analysis — Step 5"
subtitle: "Embedding-Weighted Transfer Learning (Model 2)"
date: today
engine: knitr
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
    theme: cosmo
    fig-width: 10
    fig-height: 6
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
if (!file.exists("_targets_embedding")) setwd("..")
library(targets); library(dplyr); library(ggplot2); library(knitr)

cfg_emb       <- tar_read(cfg_emb,          store = "_targets_embedding")
emb_weights   <- tar_read(emb_weights,       store = "_targets_embedding")
emb_tl_models <- tar_read(emb_tl_models,    store = "_targets_embedding")
emb_tl_maps   <- tar_read(emb_tl_maps,      store = "_targets_embedding")
```

## What this step does

Model 2 replaces the Wadoux random forest domain classifier (Step 4) with
**embedding cosine similarity** derived from
`GOOGLE/SATELLITE_EMBEDDING/V1/ANNUAL` — a 64-dimensional foundation model
encoding of annual satellite imagery (2023–2025 average).

**Key difference from Step 4:**

| | Step 4 (Wadoux) | Step 5 (Embeddings) |
|-|-----------------|---------------------|
| Similarity measure | RF on 6 hand-crafted spectral indices | Cosine distance in 64-d learned embedding space |
| Captures | Broad spectral/topo character | Full phenological + textural + spectral context |
| Needs local data for Stage A? | Yes — domain classifier trains on global vs local labels | No — similarity is computed directly from AOI mean embedding |
| Weight interpretation | P(local \| covariates) / P(global \| covariates) | sim^α, α = 5 |

Stages B–D (weighted global RF, bias correction, bootstrap CI) are
identical to Step 4.

## Embedding similarity weights

### Weight distribution

```{r weight-dist, fig.height=4}
print(emb_tl_maps$weights)
```

### Cosine similarity summary

```{r sim-table}
emb_weights |>
  summarise(
    `N cores`    = n(),
    `Min sim`    = round(min(cosine_sim),    3),
    `Median sim` = round(median(cosine_sim), 3),
    `Max sim`    = round(max(cosine_sim),    3),
    `Weight > 1` = sum(weight > 1),
    `Weight > 2` = sum(weight > 2),
    `Weight > 5` = sum(weight > 5)
  ) |>
  kable(caption = "Embedding cosine similarity to AOI mean vector (α = 5 sharpening)")
```

Cores with **cosine similarity > 0.95** are considered strongly similar to
the local AOI. The sharpening exponent (α = 5) down-weights moderately
similar cores aggressively, so only the closest analogues dominate the
global RF.

### Covariate similarity heatmap

```{r heatmap, fig.height=9}
print(emb_tl_maps$heatmap)
```

## Model performance

```{r cv-table}
summary_df <- emb_tl_maps$summary

summary_df |>
  mutate(
    depth_interval = c("0–15 cm", "15–30 cm", "30–50 cm", "50–100 cm")[
      match(depth_cm, c(7.5, 22.5, 40, 75))
    ],
    bias_correction = sprintf("%+.3f", bias_correction),
    cv_r2_tl        = ifelse(is.na(cv_r2_tl),     "—", sprintf("%.3f", cv_r2_tl)),
    cv_r2_global    = ifelse(is.na(cv_r2_global),  "—", sprintf("%.3f", cv_r2_global)),
    cv_rmse_tl      = ifelse(is.na(cv_rmse_tl),    "—", sprintf("%.4f", cv_rmse_tl)),
    cv_rmse_global  = ifelse(is.na(cv_rmse_global),"—", sprintf("%.4f", cv_rmse_global))
  ) |>
  select(
    `Depth`       = depth_interval,
    `N global`    = n_global,
    `N local`     = n_local,
    `Covariates`  = n_covariates,
    `Bias (kg/m²)`= bias_correction,
    `Bias SE`     = bias_se,
    `TL R²`       = cv_r2_tl,
    `Global R²`   = cv_r2_global,
    `TL RMSE`     = cv_rmse_tl,
    `Global RMSE` = cv_rmse_global
  ) |>
  kable(align = c("l", rep("r", 9)),
        caption = "LOCO CV: embedding-weighted TL vs uncorrected global model")
```

### Cross-validation comparison

```{r cv-plot}
print(emb_tl_maps$validation)
```

## Spatial predictions

### Global prior vs transfer-corrected

```{r maps, fig.height=8}
print(emb_tl_maps$maps)
```

## Step 4 vs Step 5 comparison

The table below summarises the key difference between the two transfer
learning approaches. Lower RMSE and less-negative R² indicate better
bias-corrected predictions for the local site.

```{r comparison}
step4 <- tryCatch(
  tar_read(tl_maps, store = "_targets_transfer")$summary,
  error = function(e) NULL
)

if (!is.null(step4)) {
  step4_slim <- step4 |>
    select(depth_cm, cv_rmse_tl, cv_r2_tl) |>
    rename(wadoux_rmse = cv_rmse_tl, wadoux_r2 = cv_r2_tl)

  summary_df |>
    select(depth_cm, cv_rmse_tl, cv_r2_tl) |>
    rename(emb_rmse = cv_rmse_tl, emb_r2 = cv_r2_tl) |>
    left_join(step4_slim, by = "depth_cm") |>
    mutate(
      rmse_delta = round(emb_rmse - wadoux_rmse, 4),
      r2_delta   = round(emb_r2   - wadoux_r2,   3),
      better     = ifelse(rmse_delta < 0, "Embedding", "Wadoux")
    ) |>
    kable(
      col.names = c("Depth cm", "Emb RMSE", "Emb R²",
                    "Wadoux RMSE", "Wadoux R²",
                    "ΔRMSE", "ΔR²", "Better"),
      caption = "Step 4 (Wadoux) vs Step 5 (Embedding) LOCO CV"
    )
} else {
  cat("Step 4 store not found — run _targets_transfer pipeline to enable comparison.")
}
```

## How to use these outputs

The embedding-weighted rasters are written to `outputs/transfer/` with the
same 4-band structure as Step 4 (Global Prior, Transfer Final, Local Only,
Difference per depth). Use **Transfer Final** for VM0033 reporting.

To generate 90% prediction intervals:

```r
m  <- emb_tl_models$models[["7.5"]]
se <- sqrt(m$bias_se^2 + m$residual_sd^2)
z  <- qnorm(0.95)   # one-sided 95% = two-sided 90% CI
# For a pixel with transfer_final prediction p:
lower <- p - z * se
upper <- p + z * se
```

**Next steps:**

- Compare Step 4 and Step 5 spatial patterns — do they agree on high/low
  carbon zones, or diverge in specific habitat types?
- If embeddings improve LOCO CV RMSE, adopt as primary method for VM0033.
- Consider Option 2 (cluster-based pixel-wise weights) if the AOI shows
  strong spatial heterogeneity in the embedding heatmap.
